\pkg{cubble}: An R Package for Structuring Spatio-temporal Data

### 2021-12-06 main.Rmd
interactive graphic section finishes

**Our new data structure for spatio-temporal data**\
This paper describes the implementation of a new spatio-temporal data structure: \pkg{cubble}. \pkg{cubble} implements a relational data structure that uses two forms to manage the switch between spatial and temporal dimension. With this structure, users can manipulate the spatial or temporal dimension separately, while leaves the linking of two dimensions to \pkg{cubble\pkg{. The software is available from the Comprehensive R Archive Network (CRAN) at [CRAN link]. \newline

@@

Interactive graphics give users the chance to manipulate the graphes for additional information. This is especially useful for spatio-temporal data since users can zoom or pan the map to see more details; use tooltips or popups to query more information about a graphic element; or explore points in multiple facets through highlight and cross-linking.

Current general purpose R implementations on interactive graphics include: \pkg{plotly} [@plotly], \pkg{leaflet} [@leaflet] that uses the underlying javascript library to realise various interactive actions; \pkg{ggiraph} [@ggiraph] implements tooltip, self-linking, and customised actions specified through \\proglang{javascripts}; \pkg{crosstalk} [@crosstalk] implements brushed linking between htmlwidgets. There's also more R implements that produces leaflet extension, i.e. \pkg{tmap} [@tmap], \pkg{mapview} [@mapview], and \pkg{leafpop} [@leafpop].

While many graphic implementations present worked examples to illustrate the usage of the package, few documents the design framework that transforms the raw data step-by-step to the final view on the screen. This step is called data pipeline and some early work in building a data pipeline includes @buja1988elements, @buja1996interactive, and  @sutherland2000orca. @wickham2009plumbing argues that *this pipeline must be presented, in some form, in all graphic software*. For interactive graphics, the pipeline also needs to include how this workflow is updated to reflect a captured user action. Some good examples of this includes [@xie2014reactive]. maybe one sentence on @cheng2016enabling.  In @wickham2009plumbing, they hope to implement a workflow where subseting of the data can be moved to a later stage. This would enable the plot to display different subsets of the data. @xie2014reactive describes the machanism of category linking with multiple sources through a key variable.

Here we will describe brushed linking between the two forms in cubble. With spatial temporal data, a common type of user action is to link between the time series plotted with the site on the map. Figure \ref{fig:interactive-workflow} shows how does interactivities work with graphics created by two forms in a cubble.  With cubble, the time series plot and site map can be created with the long and nested form separately. The unique identifier shared by both the nested and long form in a cubble is the `key` variable. Once a user action is captured, the system will look for the corresponding key variable, which will highlight the observations in both forms. Then graphics will be updated with the highlighted observations.

@@

linking machenism in crosstalk and leaflet

>>>>

**Our new data structure for spatio-temporal data**
This paper describes the implementation of a new spatio-temporal data structure: \pkg{cubble}. \pkg{cubble} implements a relational data structure that uses two forms to manage the switch between spatial and temporal dimension. With this structure, users can manipulate the spatial or temporal dimension separately, while leaves the linking of two dimensions to \pkg{cubble\pkg{. The software is available from the Comprehensive R Archive Network (CRAN) at [CRAN link]. \newline

@@

Interactive graphics can listen to users' actions on the plot to provide additional information that facilitates data exploration. This is a useful technique for spatio-temporal data since users can zoom or pan the map to view the local and global structure of the map; use tooltips or popups to query more information about a graphic element; or highlight points to explore its linked views in other plots. In the R community, many implementations have been developed to connect \proglang{R} to \proglang{javascript} to create interactive graphics. In relation to spatio, temporal, and spatio-temporal data, the general purpose packages \pkg{plotly} [@plotly] and \pkg{leaflet} [@leaflet] realise various interactive actions through their corresponding javascript libraries. \pkg{crosstalk} [@crosstalk] and \pkg{tsibbletalk} [@tsibbletalk] implement brushed linking between htmlwidgets. \pkg{ggiraph} [@ggiraph] enables tooltip, self-linking, and customised actions specified through its own \proglang{javascript}.

While many graphic implementations present worked examples to illustrate the usage of the package, few documents the underlying pipeline that transforms the raw data step-by-step to the final view on the screen. There have been some early work in building the data pipeline for (interactive) graphics [@buja1988elements; @buja1996interactive;  @sutherland2000orca] and more recent discussions include @wickham2009plumbing, @xie2014reactive, and @cheng2016enabling.

Figure \ref{fig:interactive-workflow} shows how cross-linking works with the two forms in a cubble. The data pipeline flows from the *Augmented Data* to *Transformation*, and then *Subset*, to finally, *Plot*. This data pipeline, proposed by @wickham2009plumbing, places the *Subset* stage as the last step before *Plot* so that each plot can be made separately from different subsets. This design fits naturally with cubble where the spatial map is solely made from the nested form and the time series plot can be created from the long form. When a user action is captured either from the map or the time series, the relevant row will be highlighted in the data and all the observations that shares the same `key` variable (`id`), in both forms, will then be highlighted. This type of linking is called categorical linking [@xie2014reactive], which is a one-to-many linking through highlighting observations within the same category.

@@

### 2021-12-06 main.tex
interactive graphic section finishes

\textbf{Our new data structure for spatio-temporal data}\\
This paper describes the implementation of a new spatio-temporal data
structure: \pkg{cubble}. \pkg{cubble} implements a relational data
structure that uses two forms to manage the switch between spatial and
temporal dimension. With this structure, users can manipulate the
spatial or temporal dimension separately, while leaves the linking of
two dimensions to \textbackslash pkg\{cubble\textbackslash pkg\{. The
software is available from the Comprehensive R Archive Network (CRAN) at
{[}CRAN link{]}. \newline

@@

Interactive graphics give users the chance to manipulate the graphes for
additional information. This is especially useful for spatio-temporal
data since users can zoom or pan the map to see different areas; use
tooltips or popups to query more detailed information about a particular
point; or explore points in multiple facets through highlight and
cross-linking.

Current general purpose R implementation on iterative graphics includes:
\pkg{plotly} \citep{plotly}, \pkg{leaflet} \citep{leaflet}, and
\pkg{mapdeck} \citep{mapdeck} uses the underlying javascript library
\pkg{plotly.js}, \pkg{leaflet.js}, and \pkg{mapbox.js}; \pkg{ggiraph}
\citep{ggiraph} for {[}\ldots{]}; \pkg{crosstalk} \citep{crosstalk} has
developed a cross linking between html widgets. There's also more R
implement that produces leaflet extension, i.e.~\pkg{tmap} \citep{tmap},
\pkg{mapview} \citep{mapview}, and \pkg{leafpop} \citep{leafpop}.

While many graphic implementations present the design framework and
presents worked examples, few demonstrates how the design framework
transforms the raw data step-by-step to the final view on the screen.
This step is called data pipeline and some early work in building a data
pipeline includes \citet{buja1988elements}, \citet{buja1996interactive},
and \citet{sutherland2000orca}. \citet{wickham2009plumbing} argues that
\emph{this pipeline must be presented, in some form, in all graphic
software}. For interactive graphics, the pipeline also needs to include
how this workflow is updated to reflect a captured user action. Some
good examples of this includes \citep{xie2014reactive}. maybe one
sentence on \citet{cheng2016enabling}. In \citet{wickham2009plumbing},
they hope to implement a workflow where subseting of the data can be
moved to a later stage. This would enable the plot to display different
subsets of the data. \citet{xie2014reactive} describes the machanism of
category linking with multiple sources through a key variable.

Here we will describe brushed linking between the two forms in cubble.
With spatial temporal data, a common type of user action is to link
between the time series plotted with the site on the map. Figure
\ref{fig:interactive-workflow} shows how does interactivities work with
graphics created by two forms in a cubble. With cubble, the time series
plot and site map can be created with the long and nested form
separately. The unique identifier shared by both the nested and long
form in a cubble is the \texttt{key} variable. Once a user action is
captured, the system will look for the corresponding key variable, which
will highlight the observations in both forms. Then graphics will be
updated with the highlighted observations.

>>>>

\textbf{Our new data structure for spatio-temporal data} This paper
describes the implementation of a new spatio-temporal data structure:
\pkg{cubble}. \pkg{cubble} implements a relational data structure that
uses two forms to manage the switch between spatial and temporal
dimension. With this structure, users can manipulate the spatial or
temporal dimension separately, while leaves the linking of two
dimensions to \textbackslash pkg\{cubble\textbackslash pkg\{. The
software is available from the Comprehensive R Archive Network (CRAN) at
{[}CRAN link{]}. \newline

@@

Interactive graphics can listen to users' actions on the plot to provide
additional information that facilitates data exploration. This is a
useful technique for spatio-temporal data since users can zoom or pan
the map to view the local and global structure of the map; use tooltips
or popups to query more information about a graphic element; or
highlight points to explore its linked views in other plots. In the R
community, many implementations have been developed to connect
\proglang{R} to \proglang{javascript} to create interactive graphics. In
relation to spatio, temporal, and spatio-temporal data, the general
purpose packages \pkg{plotly} \citep{plotly} and \pkg{leaflet}
\citep{leaflet} realise various interactive actions through their
corresponding javascript libraries. \pkg{crosstalk} \citep{crosstalk}
and \pkg{tsibbletalk} \citep{tsibbletalk} implement brushed linking
between htmlwidgets. \pkg{ggiraph} \citep{ggiraph} enables tooltip,
self-linking, and customised actions specified through its own
\proglang{javascript}.

While many graphic implementations present worked examples to illustrate
the usage of the package, few documents the underlying pipeline that
transforms the raw data step-by-step to the final view on the screen.
There have been some early work in building the data pipeline for
(interactive) graphics
\citep{buja1988elements, buja1996interactive, sutherland2000orca} and
more recent discussions include \citet{wickham2009plumbing},
\citet{xie2014reactive}, and \citet{cheng2016enabling}.

Figure \ref{fig:interactive-workflow} shows how cross-linking works with
the two forms in a cubble. The data pipeline flows from the
\emph{Augmented Data} to \emph{Transformation}, and then \emph{Subset},
to finally, \emph{Plot}. This data pipeline, proposed by
\citet{wickham2009plumbing}, places the \emph{Subset} stage as the last
step before \emph{Plot} so that each plot can be made separately from
different subsets. This design fits naturally with cubble where the
spatial map is solely made from the nested form and the time series plot
can be created from the long form. When a user action is captured either
from the map or the time series, the relevant row will be highlighted in
the data and all the observations that shares the same \texttt{key}
variable (\texttt{id}), in both forms, will then be highlighted. This
type of linking is called categorical linking \citep{xie2014reactive},
which is a one-to-many linking through highlighting observations within
the same category.
